---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
```


* * *

## Part 1: Data

Obviouly, the study is an observational study.  From this study, we cannot make causal conclusions on any two fractors.
However, we can make hypotheses on associations, and verify them with analysis.


* * *

## Part 2: Research questions

**Research quesion 1:**

If you random select one sample who had ever tested HIV, what's the probability of the sample being male?

**Research quesion 2:**

Is there any difference in HIV test coverage among males and females?
Try to find convincing evidence for your conclusion.


**Research quesion 3:**

Is there any association between education level and income level?

* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**

The question is a conditional probability problem.  First we need to know the distribution.

```{r}

q1.tbl <- table(sex = brfss2013$sex, hiv=brfss2013$hivtst6)
q1.tbl

```

P = 54829/(54829+77345) = 0.4148244, about 41.5%.


**Research quesion 2:**

From the data set, we extract the following two fields:

1. sex: Respondents Sex
2. hivtst6: Ever Tested HIV

First we divide the data set and count the number according to the two variables.


```{r}
sexHiv.tbl <- table(sex = brfss2013$sex, hiv = brfss2013$hivtst6)
sexHiv.tbl
```

From the results, we can calculate:

1. Sample Size

  i. Number of Males = 54829+122787 = 177616
  ii. Number of Females = 77345+182221 = 259566
  iii. Total sample size = 177616+259566 = 437182
  
2. Among all population, 

  i. percent(males) = 177616/437182 = 0.4062747
  ii. percent(females) = 259566/437182 = 0.5937253
   
3. Coverage of HIV test among males =  54829/177616 = 0.308694
4. Coverage of HIV test among females = 77345/259566 = 0.2979782
5. Total coverage of HIV test among all sex = (54829+77345)/437182 = 0.3023318


The difference of coverage rate is `0.308694 - 0.2979782 = 0.0107158`, 
namely 1.07%.  The difference is very little, but we cannot make sure whether there is a difference in HIV test coverage among males or females.

Here we make a hyperthesis with two claims:

H0: Independence model. The variables *sex* and *hiv* are independent.  They have no relationship.  The difference rate 1.07%, is due to chance.
HA: Alternative model.  The variables *sex* and *hiv* are not independent.

We shall do a simluation for HIV test coverage among the same population.


```{r}
# HIV test coverage of the whole sample
hivTestCoverRate <- 0.3023318

maleNumber <- 177616
femaleNumber <- 259566
totalNumber <- 437182

# assign sex to the population
sexHivSim.sexList <- c(rep("Male", maleNumber), rep("Female", femaleNumber))

# the function defines how to simluate the difference of HIV test coverage rate among males and females.
simulate_diff_rate <- function(sexList, sampleSize, hivTestCov) {
  hivList <- sample(c(TRUE, FALSE), totalNumber, replace=TRUE, prob=c(hivTestCov, 1.0 - hivTestCov))
  df <- data.frame(sex = sexHivSim.sexList, hiv = hivList)
  maleHiv <- df %>% filter(sex=="Male", hiv==TRUE) %>% summarise(total=n())
  femaleHiv <- df %>% filter(sex=="Female", hiv==TRUE) %>% summarise(total=n())

  # calculate the difference of HIV test coverage between male and female.
  maleCoverage <- as.double(maleHiv$total)/as.double(maleNumber)
  femaleCoverage <- as.double(femaleHiv$total)/as.double(femaleNumber)
  maleCoverage - femaleCoverage
}

simulate_diff_rate(sexHivSim.sexList, totalNumber, hivTestCoverRate)
```

The code above is one simulation.  In order to know the distribution of coverage difference, we will do 1000
simulations. Here is the code:

```{r}
# it takes about 10 seconds to run the simulation
sexHivSim.func <- function(x) simulate_diff_rate(sexHivSim.sexList, totalNumber, hivTestCoverRate)
sexHivSim.diffDistribution <- data.frame(diff= sapply(X=1:100,FUN=sexHivSim.func))

# format the data in precision 3, and draw a stack dot plot
stripchart(round(sexHivSim.diffDistribution, 3), method="stack", pch=1, ylim=c(0,100))
```

From the plot, we can tell the difference values are of a normal distribution.  So we can use `pnorm` to calculate the location of `0.0107` in the distribution:

```{r}
# calculate the mean and sd
sexHivSim.res <- sexHivSim.diffDistribution %>% summarise(mean=mean(diff), median=median(diff), sd=sd(diff), iqr=IQR(diff))

sexHivSim.res

# find the location of 0.0107
pnorm(0.0107, mean=sexHivSim.res$mean, sd=sexHivSim.res$sd)
```

The result is `1`.  It means that no values bigger than 0.0107 in the results simulated. In other words, the difference of at least 1.07% due to chance alone would
not happen.  So we prove H0 false, and HA true.

The conclusion is there is an association between sex and HIV test coverage (rate).

**Research quesion 3:**

To illustrate the percentage of different income levels for samples with different education levels, we draw a bar plot of income level percent by education level.  check the following code:

```{r}
eduIncome.tbl <- table(brfss2013$educa, brfss2013$income2)
eduIncome.xnames <- c("No/Little education", "Elementary", "Some high school", "High school graduate", "Some college techincal school", "College graduate")
eduIncome.xnames2 <- c("No/Little", "Elementary", "High school", "High school graduate", "College/techincal school", "College graduate")
eduIncome.colors <- c("#e6e6e6", "#d7d7d7", "#c8c8c8", "#b7b7b7", "#a4a4a4", "#8e8e8e", "#737373", "#4d4d4d")
barplot(t(prop.table(eduIncome.tbl, 1)), main="Distribution of Income Level by Education Level", xlab="education level", ylab="income percent", col=eduIncome.colors, legend=TRUE, names.arg = eduIncome.xnames2)

```

From the plot, we know two things clearly:
1. strictly, the percent of "Less than $10,000" decreases as the education level increases
2. generally, the percent of "$75,000 or more" increases as the education level increases

We have similar findings with mosiac plot (to make it look pretty, we convert the value to shorter ones):

```{r}
convertIncome <- function(income_level) {
  case_when(
         income_level == "Less than $10,000" ~ 1,
         income_level == "Less than $15,000" ~ 2,
         income_level == "Less than $20,000" ~ 3,
         income_level == "Less than $25,000" ~ 4,
         income_level == "Less than $35,000" ~ 5,
         income_level == "Less than $50,000" ~ 6,
         income_level == "Less than $75,000" ~ 7, 
         income_level == "$75,000 or more" ~ 8)
}
convertEducation <- function(edu_level) {
  case_when(
    edu_level == "Never attended school or only kindergarten" ~ 1,
    edu_level == "Grades 1 through 8 (Elementary)" ~ 2, 
    edu_level == "Grades 9 though 11 (Some high school)" ~ 3,
    edu_level == "Grade 12 or GED (High school graduate)" ~ 4,
    edu_level == "College 1 year to 3 years (Some college or technical school)" ~ 5,
    edu_level == "College 4 years or more (College graduate)" ~ 6)
}

eduIncome.numTbl <- table(edu_level=convertEducation(brfss2013$educa), income_level=convertIncome(brfss2013$income2))

mosaicplot(eduIncome.numTbl, 
           main="Mosaic Plot on Relations of Education level and Income level", 
           xlab="Education Level", ylab="Income Level", 
           color = eduIncome.colors)
```

From the mosaic plot, we can also find out that the population with higer education levels tends to be larger (the width increases from left to right).

However, we cannot conclude the increase of income level is associated with higher education level.
Here, analyze the data structure of `eduIncome.tbl`, we know that it's a [two-way table](http://stattrek.com/statistics/two-way-table.aspx "two way table") because there are two
categorical variable: `education level` and `income level`.  Usually we use Chi-Square test to assess the associations. 
Here is the code:

```{r}
chisq.test(eduIncome.tbl)
```

We see that p-value < 2.2e-16 < 0.05, which is statistically significant.  So we can conclude that
`there is a positive association between education level and income level`.

## References:

1. http://sites.stat.psu.edu/~ajw13/stat200_notes/12_assoc/10_assoc_print.htm
2. http://www.r-tutor.com/elementary-statistics/goodness-fit/chi-squared-test-independence
